---
title: "Load projections all bay segments"
format: 
  html:
    toc: true
    code-fold: true
    code-tools: 
      source: https://github.com/tbep-tech/otbasscap-data/blob/main/docs/load_increase_allseg.qmd
editor: source
lightbox: true

execute: 
  warning: false
  message: false
  echo: true
  fig-height: 8
  fig-width: 8
---

```{r}
library(tidyverse)
library(sf)

load(file = here::here('data-raw/mosdat.RData'))
load(file = here::here('data-raw/mohydat.RData'))
load(file = here::here('data-raw/tnanndat.RData'))
load(file = here::here('data/allsubbas.RData'))

segs <- c('Old Tampa Bay', 'Hillsborough Bay', 'Middle Tampa Bay')

# monthly data
lddatmo <- mosdat |> 
  filter(bay_segment %in% segs) |> 
  filter(year >= 2000 & year <= 2021) |> 
  select(-bod_load) |> 
  summarise(
    tp_load = sum(tp_load), 
    tss_load = sum(tss_load),
    .by = c('year', 'month', 'bay_segment')
  ) |> 
  left_join(mohydat, by = c('year', 'month', 'bay_segment')) |> 
  pivot_longer(tp_load:hy_load_106_m3_mo, names_to = 'var', values_to = 'val')
npstnlddatmo <- mosdat |> 
  filter(bay_segment %in% segs) |> 
  filter(year >= 2000 & year <= 2021) |> 
  filter(source == 'NPS') |>
  mutate(var = c('tn_load')) |> 
  select(year, month, bay_segment, var, val = tn_load)
lddatmo <- lddatmo |> 
  bind_rows(npstnlddatmo) |> 
  mutate(
    bay_segment = factor(bay_segment, levels = segs),
    var = factor(var, levels = c('tn_load', 'tp_load', 'tss_load', 'hy_load_106_m3_mo'), 
                 labels = c('TN (NPS)', 'TP' ,'TSS', 'HY'))
  ) |>   
  arrange(year, month, bay_segment, var)

# annual data
lddatann <- lddatmo |> 
  summarise(
    val = sum(val), 
    .by = c('year', 'bay_segment', 'var')
  ) 
```

This document shows a projection of estimated load change to 2040 based on annual loading estimates from 2000 to 2021 for all bay segments, except LTB. 

## Estimating projections

A regression fit is first made from 2000 to 2021 for each parameter and bay segment, then projected to 2040.  Note that this is only done for NPS for TN.  All other parameters use the combined load sources for the projection.

```{r}
# regression model of tn_load by year
mod <- lm(val ~ year*bay_segment*var, data = lddatann)

# get predictions for years 2000:2040
preds <- data.frame(year = 2000:2040) |> 
  crossing(
    bay_segment = factor(segs, levels = segs), 
    var = factor(levels(lddatann$var), levels = levels(lddatann$var))
  )
preds <- preds |> 
  mutate(
    val = predict(mod, preds)
  ) 

# plot tn_load as line plot by year
# add regression predictions
p <- ggplot(lddatann, aes(x = year, y = val)) +
  geom_line() +
  geom_line(data = preds, linetype = 'dashed') +
  geom_smooth(method = 'lm', se = F, formula = y ~ x) + 
  facet_wrap(var~bay_segment, ncol = 3, scales = 'free_y') + 
  labs(title = 'Loads by bay segment',
       x = 'Year',
       y = 'Load (tons / yr)',
       caption = 'note: HY (hydrology) is million gallons per year'
  ) +
  theme_minimal()
p
```

The percent change is calculated as the average value from the regression fit for 2000 to 2021 to 2040.

```{r}
# estimate means from 2000 to 2021 and 2040
means <- preds |> 
  summarise(
    mean_load_2000_2021 = mean(val[year >= 2000 & year <= 2021]),
    mean_load_2040 = mean(val[year == 2040]), 
    .by = c(bay_segment, var)
  )

# take percent change from 2000 to 2021 to 2040
means <- means |> 
  mutate(
    percent_change = (mean_load_2040 - mean_load_2000_2021) / mean_load_2000_2021 * 100, 
    .by = c(bay_segment, var)
  )
means
```

The estimated changes can then be projected onto to observed load values for 2000 to 2021.  Note that some have negative projections due to extrapolation of the regression to values less than zero.

```{r}
lddatmochg <- lddatmo |> 
  left_join(means, by = c('bay_segment', 'var')) |> 
  select(-mean_load_2000_2021, -mean_load_2040) |> 
  mutate(
    valchg = val + (val * (percent_change / 100)), 
    date = make_date(year, month, '1')
  )

p <- ggplot(lddatmochg, aes(x = date)) +
  geom_line(aes(y = val, color = 'original')) +
  geom_line(aes(y = valchg, color = 'projected')) +
  facet_wrap(var~bay_segment, ncol = 3, scales = 'free_y') + 
  labs(title = 'Loads by bay segment',
       x = NULL,
       y = 'Load (tons / mo)', 
       color = NULL,
       caption = 'note: HY (hydrology) is million gallons per month'
  ) +
  theme_minimal() + 
  theme(legend.position = 'top')
p
```

Verify the percent change is correct.  This should be the same as the `means` object.

```{r}
lddatmochg |> 
  summarize(
    val = sum(val),
    valchg = sum(valchg), 
    .by = c(bay_segment, var)
  ) |> 
  mutate(
    percent_change =  (valchg - val) / val * 100
  )
```

The last step is to add the NPS change for TN to the total TN load and assign the projected values to the next twenty years. Projected values are also floored at zero.

```{r}
tnnpsmochg <- lddatmochg |> 
  filter(var == 'TN (NPS)') |> 
  select(year, month, bay_segment, valchg)
tnmodat <- mosdat |> 
  filter(year >= 2000 & year <= 2021) |> 
  filter(bay_segment %in% segs) |> 
  select(-tp_load, -tp_load, -tss_load, -bod_load) |> 
  left_join(tnnpsmochg, by = c('year', 'month', 'bay_segment')) |> 
  mutate(
    valchg = case_when(
      source == 'NPS' ~ valchg, 
      T ~ tn_load
    ),
    var = 'TN',
    bay_segment = factor(bay_segment, levels = segs)
  ) |> 
  summarise(
    val = sum(tn_load),
    valchg = sum(valchg),
    .by = c(year, month, bay_segment, var)
  )
projected <- lddatmochg |> 
  filter(var != 'TN (NPS)') |> 
  select(year, month, bay_segment, var, val, valchg) |> 
  bind_rows(tnmodat) |> 
  mutate(var = factor(var, levels = c('TN', 'TP', 'TSS', 'HY'))) |> 
  pivot_longer(c(val, valchg), names_to = 'period', values_to = 'val') |> 
  mutate(
    year = case_when(
      period == 'valchg' ~ year + 22, 
      T ~ year
    ),
    date = make_date(year, month, '1'), 
    period = factor(period, levels = c('val', 'valchg'), labels = c('observed', 'projected')),
    val = pmax(val, 0)
  ) |> 
  arrange(date, bay_segment, var)

p <- ggplot(projected, aes(x = date, y = val, color = period)) +
  geom_line() +
  facet_wrap(var~bay_segment, ncol = 3, scales = 'free_y') + 
  labs(title = 'Loads by bay segment',
       x = NULL,
       y = 'Load (tons / mo)', 
       color = NULL,
       caption = 'note: HY (hydrology) is million gallons per month)'
  ) +
  theme_minimal() + 
  theme(legend.position = 'top')
p
```

## Assigning to model segments

There are three model segments for OTB, two for HB, and three for MTB.  The loads are apportioned in to equal amounts based on the number of model segments per segment, except OTB which uses amounts described [here](https://tbep-tech.github.io/otbasscap-data/otb_subbasin_loads.html).  The previous link assigns values to NW, NE, CW, CE, SW, and SE sub-segments as sixths, which are assigned to model segments 1, 2, and 3 as NW/NE, CW/CE, and SW/SE.  Here we simply apportion using the relative areas that drain to each OTB subsegment, whereas the previous example used a hydrologic basin dataset from 2017-2021 that had gaged and ungaged loading estimates.  

```{r}
otbprop <- allsubbas |> 
  mutate(
    acres = units::set_units(st_area(allsubbas), 'acres'), 
    modelseg = case_when(
      subsegment %in% c('NW', 'NE') ~ 1, 
      subsegment %in% c('CW', 'CE') ~ 2, 
      subsegment %in% c('SW', 'SE') ~ 3
    )
  ) |> 
  st_set_geometry(NULL) |> 
  group_by(modelseg) |> 
  summarise(
    acres = sum(acres)
  ) |> 
  mutate(
    bay_segment = 'Old Tampa Bay',
    prop = as.numeric(acres / sum(acres))
  ) |> 
  select(-acres)

allprop <- tibble(
    modelseg  = c(4:8),
    bay_segment = c(rep('Hillsborough Bay', 2), rep('Middle Tampa Bay', 3)),
    prop = c(rep(0.5, 2), rep(1/3, 3))
  ) |> 
  bind_rows(otbprop) |> 
  mutate(bay_segment = factor(bay_segment, levels = segs)) |> 
  arrange(bay_segment)
allprop
```

The `allprop` object is joined to the observed and projected monthly loads by bay segment and then the loads are multiplied by the proportions.

```{r}
out <- allprop |> 
  left_join(projected, by = 'bay_segment', relationship = c('many-to-many')) |> 
  mutate(val = val * prop)
out
```

Now verify that the totals are the same between the model segment assigned dataset and the original.

```{r}
summarise(out, val = sum(val), .by = c(bay_segment, var))
summarise(projected, val = sum(val), .by = c(bay_segment, var))
```

```{r}
write.csv(out, here::here('data-raw/boxmodproject.csv'), row.names = F)
```

